---
title: "Notebook to reproduce Figure 2d & 2e"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
author: "Joan Kant"
params:
    output_dir: "output/VisiumHD"
    sample_id:  "Gaiti_Yiyan__6425_cortex_1_A1"
    segmentation_masks_path: "data/VisiumHD/processed/6425_cortex_1_A1__masks.json"
    seurat_obj_path: "data/VisiumHD/processed/6425_cortex_1_A1.rds"
    output_dir: "output/VisiumHD"
    k_neighbors: 30
    pseudobulk_scores_path: "data/VisiumHD/processed/10k/6425_combined_pseudobulk_scores_k30.rds"
    n_iter: 10000
    lower_limit: 0.10
    upper_limit: 0.90
    scale_factors_path: "/cluster/projects/gaitigroup/Data/GBM_VisiumHD/20240708_LH00244_0121_B22MN7GLT3_Gaiti_Yiyan_Visium/6425_cortex_1_A1/outs/binned_outputs/square_002um/spatial/scalefactors_json.json"
    seurat_obj_subset_path: "data/VisiumHD/processed/10k/6425_combined_seurat_obj_meanshortest_k30.rds"
    markers_path: "data/VisiumHD/processed/6425_cortex_1_A1_markers_by_celltype.rds"

output: html_document
---

```{r, rmarkdown-setup, include=FALSE}
knitr::opts_knit$set(
    root.dir = here::here()
)
knitr::opts_chunk$set(
    fig.align = "center", fig.width = 12, warning = FALSE
)
pacman::p_load(GaitiLabUtils, GBMutils, glue, data.table, tidyverse, stringr, sf, rjson, Seurat)
logr <- init_logging()


params <- list()
params$sample_id <- "Gaiti_Yiyan__6425_cortex_1_A1"
params$segmentation_masks_path <- glue("data/VisiumHD/processed/{params$sample_id}__masks.json")
params$seurat_obj_path <- glue("data/VisiumHD/processed/{params$sample_id}.rds")
params$output_dir <- "output/VisiumHD/20250106"
params$k_neighbors <- 30

params$pseudobulk_scores_path <- glue("data/VisiumHD/processed/10k/6425_combined_pseudobulk_scores_k{params$k_neighbors}.rds")
params$n_iter <- 10000
params$lower_limit <- 0.10
params$upper_limit <- 0.90
params$scale_factors_path <- glue("/cluster/projects/gaitigroup/Data/GBM_VisiumHD/20240708_LH00244_0121_B22MN7GLT3_Gaiti_Yiyan_Visium/{params$sample_id}/outs/binned_outputs/square_002um/spatial/scalefactors_json.json")
params$seurat_obj_subset_path <- glue("data/VisiumHD/processed/10k/6425_combined_seurat_obj_meanshortest_k{params$k_neighbors}.rds")
params$markers_path <- glue("data/VisiumHD/processed/{params$sample_id}_markers_by_celltype.rds")
```

## Setup
```{r, setup}
# Setup color palette
log_info("Load color palette...")
palette_L1 <- GBMutils::load_color_palette("Spatial_CellClass_L1")

cell_type_pair_oi <- c("Neuron", "Malignant")
malignant_signatures <- c("Differentiated-like", "Progenitor-like", "Invasive-high OPC/NPC1")
cat_dist_simplified_labels <- setNames(c("Nearest 10% malignant cells to neurons", "Furthest 10% malignant cells to neurons"), c("closest", "furthest"))

myfuns_nam <- list(close = min, far = max)
signif_level <- 0.05

# Create color palette for distance labels
distance_labels <- setNames(c(
    paste0("<", params$lower_limit * 100, "%"),
    paste0(
        params$lower_limit * 100, "% <= mean distance <= ",
        params$upper_limit * 100, "%"
    ),
    paste0(">", params$upper_limit * 100, "%")
), c("bottom", "between", "top"))

palette_dist_gradient <- setNames(RColorBrewer::brewer.pal(n = 8, name = "Blues")[c(3, 5, 8)], distance_labels)
palette_malign_dist_grad_x_neuron <- c(palette_L1["Neuron"], palette_dist_gradient)
color_gradient_simplified <- setNames(palette_dist_gradient[c(1, 3)], cat_dist_simplified_labels)
```

## Figure 2d

**Approach**
Input: two VisiumHD PT samples. 

1. For each individual sample
   1. Compute pairwise distances between Malignant cells and Neurons, i.e. results in a matrix of M x N, with M = no. malignant cells and N = no. Neurons 
   2. For each Malignant cell, compute mean distance to k-nearest Neurons, with **k=30**. Results in a vector of length M with each entry the mean distance (d). 
2. Combine the resulting vectors for the two samples. 
3. Define a lower and upper threshold, i.e. the **10th and 90th percentile**
4. Label the Malignant cells based on the defined thresholds: (1) d < d~10th, (2) d~10th <= d <= d~90th.
5. Pseudobulk by labels & normalize. 
6. Use Seurat's `AddModuleScore` to score three signatures: Differentiated-like, Progenitor-like and Invasive-high OPC/NPC1. These are the **observed scores** (s~observed)

To determine significance a permutation approach is used. 
1. Shuffle the labels of Malignant cells (see step 4 mentioned above).
2. Pseudobulk by labels & normalize. 
3. Use Seurat's `AddModuleScore` to score the three signatures (see step 6 mentioned above), these are the **shuffled scores**.
Repeat this process I times with R=10,000. 

Compute statistics: 
For each label and signature combination:
1. z-score = (s~observed - mean(S~shuffled)) / sd(S~shuffled), 
   * with s~observed, the observed score and S~shuffled, vector S with length I where each entry is a score after shuffling. 
2. p-value
   * if z-score > 0, then p-value = $$((\sum_{i=1}^{I} S~i > s~observed) + 1) / R$$
   * if z-score < 0, then p-value = $$((\sum_{i=1}^{I} S~i < s~observed) + 1) / R$$


<!-- TODO check name sample -->
Input: VisiumHD sample PT1

```{r, fig2d-prep-masks}
log_info("Load scalefactors...")
scalefactors <-
    fromJSON(file = params$scale_factors_path)

log_info("Load segmentation masks...")
gdf <- read_sf(params$segmentation_masks_path) %>% mutate(geometry_px = geometry)
# Remove coordinate systems, just want to use cartesian system
st_crs(gdf) <- NA

log_info("Load Seurat object...")
seurat_obj <- readRDS(params$seurat_obj_path)

log_info("Combine metadata and segmentation masks...")
meta_sdf <- sf::st_as_sf(
    (seurat_obj@meta.data) %>%
        left_join(gdf, by = "id")
) %>%
    # Convert to px to microns
    mutate(geometry = geometry_px * scalefactors$microns_per_pixel)
```

```{r, fig2d-neuron-malignant-continuous-distances}
log_info("Load Seurat object subset...")
seurat_obj_subset <- readRDS(params$seurat_obj_subset_path)

dist_obs_df <- seurat_obj_subset@meta.data %>%
    mutate(
        # TODO edit Seurat object later
        sample = case_when(
            sample == "PT1" ~ "Gaiti_Yiyan__6425_cortex_1_A1",
            sample == "PT3" ~ "Gaiti_Yiyan__6425_cortex_3_D1",
        ),
    ) %>%
    select(sample, id, cat_dist, mean_dist, BANKSY_snn_res.2_annot)
dist_obs_df$id <- str_split(dist_obs_df$id, "_", simplify = TRUE, n = 2)[, 2]

log_info("Combine metadata and segmentation masks...")
dist_obs_sdf <- meta_sdf %>%
    filter(BANKSY_snn_res.2_annot %in% cell_type_pair_oi) %>%
    select(id, geometry, BANKSY_snn_res.2_annot) %>%
    left_join(dist_obs_df %>%
        filter(sample == params$sample_id)) %>%
    # Convert to px to microns
    mutate(
        vis_label = factor(ifelse(BANKSY_snn_res.2_annot == "Neuron", "Neuron", cat_dist), levels = names(colors))
    )
list_val <- lapply(myfuns_nam, function(f) f(dist_obs_sdf %>% filter(BANKSY_snn_res.2_annot == "Malignant") %>% pull(mean_dist)))

p <- ggplot() +
    geom_sf(data = dist_obs_sdf %>% filter(BANKSY_snn_res.2_annot == "Malignant"), aes(fill = mean_dist), color = NA) +
    GBMutils::GBM_theme() +
    scale_fill_distiller(
        palette = "Blues", direction = -1,
        breaks = unlist(list_val),
        guide = guide_colorbar(
            override.aes = list(size = 5),
            title.position = "top",
            title = glue("Proximity to {params$k_neighbors} nearest Neurons (µm)"),
            barwidth = unit(8, "lines"), barheight = unit(0.8, "lines")
        )
    ) +
    ggnewscale::new_scale_fill() +
    geom_sf(
        data = dist_obs_sdf %>% filter(BANKSY_snn_res.2_annot == "Neuron"),
        fill = palette_L1[["Neuron"]], color = NA
    ) +
    coord_sf() +
    scale_y_reverse() +
    labs(
        x = "X (µm)", y = "Y (µm)",
        subtitle = paste0("No. cells=", prettyNum(dist_obs_sdf %>% filter(BANKSY_snn_res.2_annot %in% cell_type_pair_oi) %>% nrow(), big.mark = ","))
    ) +
    theme(aspect.ratio = 1, axis.text = element_blank(), axis.ticks = element_blank())

log_info("Save figure as PDF...")
ggsave(
    plot = p,
    filename = glue("Fig2d-{params$sample_id}_dist_gradient_k{params$k_neighbors}.pdf"),
    path = params$output_dir
)
```

## Figure 2e
Use output previously generated with `preprocessing/VisiumHD/assess_prox_vs_malign_signature.R` for visualization. 

```{r, stats}
log_info("Load pseudobulk scores...")
pseudobulk_scores_df_long <- readRDS(params$pseudobulk_scores_path) %>%
    rename(signature = variable)

# Split into observed and shuffled
pseudobulk_scores_observed_df_long <- pseudobulk_scores_df_long %>%
    filter(ix == "cat_dist") %>%
    mutate(AddModuleScore = as.numeric(AddModuleScore)) %>%
    rename(AddModuleScore_obs = AddModuleScore) %>%
    select(-ix)

pseudobulk_scores_shuffled_df_long <- pseudobulk_scores_df_long %>%
    filter(ix != "cat_dist") %>%
    mutate(AddModuleScore = as.numeric(AddModuleScore)) %>%
    rename(AddModuleScore_shuffled = AddModuleScore)

# Combine observed and shuffled dataframes
stat_df_long <- pseudobulk_scores_shuffled_df_long %>%
    left_join(pseudobulk_scores_observed_df_long)

# Convert to wide format, compute basic statistics for permutations
stat_df_wide <- stat_df_long %>%
    pivot_wider(names_from = ix, values_from = AddModuleScore_shuffled) %>%
    rowwise() %>%
    mutate(
        perm_min = min(c_across(starts_with("cat_dist_sid"))),
        perm_max = max(c_across(starts_with("cat_dist_sid"))),
        perm_mean = mean(c_across(starts_with("cat_dist_sid"))),
        perm_median = median(c_across(starts_with("cat_dist_sid"))),
        perm_sd = sd(c_across(starts_with("cat_dist_sid"))),
        # Testing hypothesis H0 < H1 (alternative = 'greater')
        count_larger = sum(c_across(starts_with("cat_dist_sid_")) > AddModuleScore_obs) + 1,
        # Testing hypothesis H0 > H1 (alternative = 'less')
        count_smaller = sum(c_across(starts_with("cat_dist_sid_")) < AddModuleScore_obs) + 1,
        z_score = (AddModuleScore_obs - perm_mean) / perm_sd,
        interaction_type = ifelse(
            z_score > 0,
            "Co-localization",
            "Avoidance"
        ),
        p_value = ifelse(interaction_type == "Co-localization",
            # Co-localization
            count_larger / params$n_iter,
            count_smaller / params$n_iter
        ),
        p_value_coloc = count_larger / params$n_iter,
        p_value_avoid = count_smaller / params$n_iter,
        cat_dist_simplified = factor(
            ifelse(cat_dist == paste0("<", params$lower_limit * 100, "%"),
                cat_dist_simplified_labels[["closest"]], cat_dist_simplified_labels[["furthest"]]
            ),
            levels = cat_dist_simplified_labels
        )
    ) %>%
    # Only interested in bottom and top category
    filter(cat_dist != distance_labels[["between"]])

# Correction for multiple testing
stat_df_wide$p_value_adj <- p.adjust(stat_df_wide$p_value, "bonferroni")

stat_df_wide <- stat_df_wide %>%
    mutate(
        is_significant = p_value < signif_level,
        is_significant_adj = p_value_adj < signif_level,
        signature = factor(str_replace_all(signature, "_", "-"), malignant_signatures)
    )
```

```{r, stats-vis-invasive}
data_bar <- stat_df_wide %>%
    filter(cat_dist_simplified == cat_dist_simplified_labels[["closest"]]) %>%
    select(signature, z_score)

p_bar <- ggplot(
    data = data_bar,
    aes(
        fill = signature,
        x = signature, y = z_score
    )
) +
    geom_bar(
        stat = "identity",
        position = "dodge",
        show.legend = FALSE,
        fill = "grey", color = "black"
    ) +
    scale_fill_manual(values = palette) +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
    GBMutils::GBM_theme() +
    labs(
        title = cat_dist_simplified_labels[["closest"]],
        subtitle =
            glue("k={params$k_neighbors} with no. iterations={prettyNum(params$n_iter, big.mark =',')}"),
        y = glue("Z-score"),
        x = "Signature"
    ) +
    coord_flip()

log_info("Save figure as PDF...")
ggsave(p_bar,
    filename = glue("Fig2e-malign_signature_vs_proximity_k{params$k_neighbors}_neurons_barplot_closest_to_neuron.pdf"),
    height = 4,
    path = params$output_dir
)

# Split into shuffled and observed
stat_shuffled_df_subset_long <- stat_df_wide %>%
    filter(
        signature == "Invasive-high OPC/NPC1",
        cat_dist_simplified == cat_dist_simplified_labels[["closest"]]
    ) %>%
    select(cat_dist_simplified, signature, starts_with("cat_dist_sid")) %>%
    reshape2::melt(variable.name = "iteration_id", value.name = "score")

stat_obs_df_subset_wide <- stat_df_wide %>%
    select(cat_dist_simplified, signature, AddModuleScore_obs, perm_mean, z_score, p_value, p_value_adj) %>%
    distinct() %>%
    filter(
        signature == "Invasive-high OPC/NPC1",
        cat_dist_simplified == cat_dist_simplified_labels[["closest"]]
    )


p_histo <- ggplot(
    data = stat_shuffled_df_subset_long,
    aes(x = score)
) +
    geom_histogram(binwidth = 1e-3, show.legend = FALSE, fill = "grey") +
    geom_vline(data = stat_obs_df_subset_wide, aes(xintercept = AddModuleScore_obs), color = "blue") +
    geom_vline(data = stat_obs_df_subset_wide, aes(xintercept = perm_mean), color = "red", linetype = "dashed") +
    geom_text(data = stat_obs_df_subset_wide, aes(
        x = (AddModuleScore_obs - AddModuleScore_obs * 0.01),
        label = glue("Observed Module score={round(AddModuleScore_obs, 5)}"), y = 300
    ), colour = "blue", angle = 90, size = 2.5) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
    GBMutils::GBM_theme() +
    labs(
        x = "Module Score", y = "No. of iterations",
        subtitle =
            glue("k={params$k_neighbors} with no. iterations={prettyNum(params$n_iter, big.mark =',')}"),
    ) +
    theme(aspect.ratio = 0.5)

log_info("Save figure as PDF...")
ggsave(p_histo,
    filename = glue("Fig2e-malign_signature_vs_proximity_k{params$k_neighbors}_neurons_invasive_only_closest_to_neuron.pdf"),
    height = 4, path = params$output_dir
)
```


```{r, manual-testing-stat, eval = FALSE}
# Check if data behind barplot matches when computing the z-scores manually
data_bar %>% head()

# Manual computation
test_df_wide <- stat_df_wide %>%
    filter(
        cat_dist_simplified == cat_dist_simplified_labels[["closest"]],
        signature == "Invasive-high OPC/NPC1"
    )

test_df_wide %>% head()

test_df_shuffled_long <- test_df_wide %>%
    select(starts_with("cat_dist_sid")) %>%
    reshape2::melt()


mean_obs <- test_df_wide %>%
    select(AddModuleScore_obs) %>%
    pull()
shuffled_scores_mean <- mean(test_df_shuffled_long$value)
shuffled_scores_sd <- sd(test_df_shuffled_long$value)
z_score <- (mean_obs - shuffled_scores_mean) / shuffled_scores_sd

# z_score should match the data_bar %>% head() output
print(z_score)
```



## DE Analysis between top and bottom 10% 

* avg_log2FC: log fold-change of the average expression between the two groups. 
  * Interpretation **Positive values indicate that the feature is more highly expressed in the first group.**
```{r, dge-analysis}
log_info("Load Seurat object...")
seurat_obj <- readRDS("data/VisiumHD/processed/6425_combined_seurat_obj_meanshortest_k30.rds")

Idents(seurat_obj) <- "cat_dist"
DefaultAssay(seurat_obj) <- "RNA"

invasive_high <- GBMutils::load_invasive_signature(type = "high")


p <- DotPlot(seurat_obj, features = intersect(invasive_high, VariableFeatures(seurat_obj)), assay = "RNA") +
    GBMutils::GBM_theme() +
    scale_colour_gradient2(midpoint = 0, low = "steelblue", high = "brown") +
    guides(
        size = guide_legend(
            direction = "vertical",
            title = "Expr. by cells (%)",
            order = 1
        ), colour = guide_colourbar(
            title = "Scaled avg. expr.",
            direction = "vertical",
            barwidth = unit(0.8, "lines"),
            barheight = unit(6, "lines"),
            title.position = "top", order = 2
        )
    ) +
    RotatedAxis() +
    labs(
        x = "Marker",
        y = "Cluster",
        subtitle = "Intersection of Invasive-high OPC/NPC1 and VariableFeatures (n=2000) "
    ) +
    theme(legend.position = "right", legend.direction = "vertical")
ggsave(p, filename = "dotplot_invasive.pdf", path = "output/testing", width = 25)


seurat_obj <- JoinLayers(seurat_obj)
de_results <- FindMarkers(seurat_obj,
    ident.1 = "<10%", ident.2 = ">90%", assay = "RNA",
    features = VariableFeatures(seurat_obj),
    # test.use = "MAST"
) %>%
    rownames_to_column("Gene")


# Thresholds
log2fc <- 1.5
p_val_adj_signif_thresh <- 0.1

cat_colors <- setNames(c("black", "blue", "darkorange"), c("NO", "YES", "Invasive signature"))


de_results_filtered <- de_results %>%
    mutate(
        diffexpressed = case_when(
            (avg_log2FC > log2fc) & (p_val_adj < p_val_adj_signif_thresh) ~ "UP",
            (avg_log2FC < -log2fc) & (p_val_adj < p_val_adj_signif_thresh) ~ "DOWN",
            .default = "NO"
        ),
        label = case_when(
            (diffexpressed != "NO") & (Gene %in% invasive_high) ~ Gene,
            .default = NA
        ),
        alpha = ifelse(is.na(label), 0.05, 1)
    ) %>%
    column_to_rownames("Gene") %>%
    mutate(diffexpressed = factor(case_when(
        diffexpressed == "NO" ~ "NO",
        rownames(.) %in% invasive_high ~ "Invasive signature",
        .default = "YES"
    ), levels = names(cat_colors)))

p <- ggplot(de_results_filtered, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
    geom_point(aes(colour = diffexpressed, alpha = alpha), size = 3.5, stroke = NA) +
    scale_colour_manual(values = cat_colors) +
    ggrepel::geom_label_repel(aes(label = label), size = 3, colour = "darkorange", na.rm = TRUE, nudge_x = 2, nudge_y = 1) +
    geom_vline(xintercept = log2fc, linetype = "dashed") +
    geom_vline(xintercept = -log2fc, linetype = "dashed") +
    geom_hline(yintercept = -log10(p_val_adj_signif_thresh), linetype = "dashed") +
    (ggthemes::theme_foundation(base_size = 14, base_family = "Helvetica")
    + theme(
            plot.title = element_text(
                face = "bold",
                size = rel(1.2), hjust = 0.5
            ),
            panel.background = element_rect(colour = NA),
            plot.background = element_rect(colour = NA),
            panel.border = element_rect(colour = NA),
            axis.title = element_text(face = "bold", size = rel(1.1)),
            axis.title.y = element_text(angle = 90, vjust = 2),
            axis.title.x = element_text(vjust = -0.2),
            axis.text = element_text(size = rel(1)),
            axis.line = element_line(colour = "black"),
            axis.ticks = element_line(),
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            legend.key = element_rect(colour = NA),
            legend.position = "bottom",
            legend.direction = "horizontal",
            legend.key.size = unit(0.2, "cm"),
            legend.spacing = unit(0, "cm"),
            legend.title = element_text(face = "plain"),
            plot.margin = unit(c(10, 5, 5, 5), "mm"),
            strip.background = element_rect(colour = "#ffffff", fill = "#ffffff"),
            strip.text = element_text(face = "bold", size = rel(1.1))
        )) +
    # ylim(c(0, 25)) +
    guides(alpha = "none") +
    xlim(c(-15, 15)) +
    xlab(bquote(~ Log[2] ~ FoldChange)) +
    ylab(bquote(~ -Log[10] ~ italic(FDR)))
ggsave(p, filename = "de_dist.pdf", path = "output/testing")
```


## Session Info
```{r, session-info, echo = FALSE}
sessionInfo()
```
